A Dynamic Intrafamily Model of Child Behavior Problems and Birth Timing

### 2019-02-21 inst/paper/paper.Rmd
Fixing paper knitting
Some tweaks to make paper knit again; removed more complex model code.

library(birthtiming) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
```

@@

```{r resultsdata, include=FALSE}
child_difficulty <- load_and_format_NLSY()
naive_tidy <- run_naive_sem()
load("../data/derived_data/naive_tidy.RData")
load("../data/derived_data/child_difficulty.RData")
```

@@

The outcome variables in this analysis are the Difficulty Composite Raw Scores for both siblings which consists of the sum of ordinal (1 to 5) responses to 11 temperament questions capturing a range of child behaviors, such as “How often do you have trouble soothing or calming your infant when he/she is crying or upset?” and “During the average day, how often does your infant get fussy and irritable?” This scale has an observed range of `r min(child_difficulty$Difficulty_1)` to `r max(child_difficulty$Difficulty_1)` (out of possible 11 to 55) for first observed children and `r min(child_difficulty$Difficulty_2)` to `r max(child_difficulty$Difficulty_2)` for second observed children. The scale is treated as a continuous measure as it is approximately normally distributed, though there is evidence of mild departures from normality for difficulty of the second child (see appendix). The predictors are maternal age at birth for each child. Mothers in the analysis subsample were `r min(child_difficulty$m_age_at_birth_1)` to `r max(child_difficulty$m_age_at_birth_1)` years of age at birth of first child and `r min(child_difficulty$m_age_at_birth_2)` to `r max(child_difficulty$m_age_at_birth_2)` for the second child
<br/><br/>
```{r knitdiagrams, include=FALSE}
knitr::knit("diagram_model.Rnw")
knitr::knit("diagram_estimates.Rnw")
```

@@

Figure \@ref(fig:diagramestimates) depicts maximum likelihood estimates of the specified structural equations obtained with `lavaan` in R. A strong negative relationship is observed between mother’s age at birth of child and perceived difficulty in early childhood. Net of time-invariant mother and family characteristics, each additional year older at childbirth corresponds to almost two points lower on the difficulty composite or an approximately 1:1 relationship in standardized metrics ($\gamma^{*}$=  `r round(naive_tidy$std.all[3],2)` and `r round(naive_tidy$std.all[4], 2)`). Additionally, we see a statistically significant (t=-2.275) relationship between the unaccounted for variation in first child difficulty and mother’s age at second birth, though this relationship is unexpectedly negative. Greater difficulty in early childhood, net of that associated with maternal age, appears associated with modestly earlier second births. As the model is just identified, tests of overall model fit cannot be conducted. For comparison the $D_{1}$ error to $T_{2}$ covariance was restricted to zero to provide fit statistics, though this restricted model is clearly misspecified given the significance of the parameter in the full model. In this model, the root mean square error of approximation is 0.0583 which indicates moderately good fit, though this may be inflated by the presence of a single degree of freedom. The restricted model chi-square is 5.731 (p=0.017), indicating worse fit than the full (saturated) model.

Some additional results notes:
1. Parity significantly predicts higher child difficulty (0.161, t=5.54). Inclusion has no effect on mother's age effects.
2. Early child health problems do not predict child difficulty or impact mother's age effects.
3. At-birth characteristics (e.g. birth weight, gestation time) do not have substantively strong effects or impacts on mother's age effects.

>>>>

# birthtiming package currently disabled while in redevelopment
library(birthtiming) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
```

@@

```{r resultsdata, include=FALSE}
library(dplyr)
library(stringr)
library(tidyr)
library(reshape2)
child_difficulty <- gen_birthtiming_data(min_children=2, max_children=10) %>%
                                        filter(!is.na(difficulty_1) & !is.na(difficulty_2)) 
naive_tidy <- run_naive_sem()
save(naive_tidy, file="../data/derived_data/naive_tidy.RData")
save(child_difficulty, file="../data/derived_data/child_difficulty.RData")
# load("../data/derived_data/naive_tidy.RData")
# load("../data/derived_data/child_difficulty.RData")
```

@@

The outcome variables in this analysis are the Difficulty Composite Raw Scores for both siblings which consists of the sum of ordinal (1 to 5) responses to 11 temperament questions capturing a range of child behaviors, such as “How often do you have trouble soothing or calming your infant when he/she is crying or upset?” and “During the average day, how often does your infant get fussy and irritable?” This scale has an observed range of `r min(child_difficulty$difficulty_1)` to `r max(child_difficulty$difficulty_1)` (out of possible 11 to 55) for first observed children and `r min(child_difficulty$difficulty_2)` to `r max(child_difficulty$difficulty_2)` for second observed children. The scale is treated as a continuous measure as it is approximately normally distributed, though there is evidence of mild departures from normality for difficulty of the second child (see appendix). The predictors are maternal age at birth for each child. Mothers in the analysis subsample were `r min(child_difficulty$m_age_at_birth_1)` to `r max(child_difficulty$m_age_at_birth_1)` years of age at birth of first child and `r min(child_difficulty$m_age_at_birth_2)` to `r max(child_difficulty$m_age_at_birth_2)` for the second child
<br/><br/>
```{r knitdiagrams, include=FALSE}
knitr::knit("diagram_model.Rnw")
knitr::knit("diagram_estimates.Rnw")
```

@@

Figure \@ref(fig:diagramestimates) depicts maximum likelihood estimates of the specified structural equations obtained with `lavaan` in R. A strong negative relationship is observed between mother’s age at birth of child and perceived difficulty in early childhood. Net of time-invariant mother and family characteristics, each additional year older at childbirth corresponds to almost two points lower on the difficulty composite or an approximately 1:1 relationship in standardized metrics ($\gamma^{*}$=  `r round(naive_tidy$std.all[3],2)` and `r round(naive_tidy$std.all[4], 2)`). Additionally, we see a statistically significant (z=`r round(naive_tidy$z[8],2)`) relationship between the unaccounted for variation in first child difficulty and mother’s age at second birth, though this relationship is unexpectedly negative. Greater difficulty in early childhood, net of that associated with maternal age, appears associated with modestly earlier second births. As the model is just identified, tests of overall model fit cannot be conducted. For comparison the $D_{1}$ error to $T_{2}$ covariance was restricted to zero to provide fit statistics, though this restricted model is clearly misspecified given the significance of the parameter in the full model. In this model, the root mean square error of approximation is 0.0583 which indicates moderately good fit, though this may be inflated by the presence of a single degree of freedom. The restricted model chi-square is 5.731 (p=0.017), indicating worse fit than the full (saturated) model.

Some additional results notes (models not included yet):

1. Parity significantly predicts higher child difficulty (0.161, t=5.54). Inclusion has no effect on mother's age effects.
2. Early child health problems do not predict child difficulty or impact mother's age effects.
3. At-birth characteristics (e.g. birth weight, gestation time) do not have substantively strong effects on child behavior or impacts on mother's age effects.
